
```{r, child = "setup.Rmd", include = FALSE}
```

## Ambiguity between columns and external variables

In selecting functions like `dplyr::select()` or `tidyr::pivot_longer()`,
you can refer to variables by name:

```{r}
mtcars %>% select(cyl, am, vs)

mtcars %>% select(mpg:disp)
```

For historical reasons, it is also possible to refer an external
vector of variable names. You get the correct result, but with a note
informing you that selecting with an external variable is brittle.

```{r}
vars <- c("cyl", "am", "vs")
result <- mtcars %>% select(vars)
```

This note will become a warning in the future, and then an error. We
have decided to deprecate this particular approach to using external
vectors because they introduce an ambiguity. Imagine that the data
frame contains a column with the same name as your external variable.

```{r}
some_df <- mtcars
some_df$vars <- 1:nrow(mtcars)
```

These are very different objects but it isn't a problem if the context
forces you to be specific about where to find `vars`:

```{r}
vars

some_df$vars
```

In a selection context however, the column wins:

```{r}
some_df %>% select(vars)
```


## Fixing the ambiguity

To make your selection code more robust and silence the message, use
`all_of()` to force the external vector:

```{r}
some_df %>% select(all_of(vars))
```

For more information or if you have comments about this, please see
the [Github issue](https://github.com/r-lib/tidyselect/issues/76)
tracking the deprecation process.
