
```{r}
#| child: "setup.Rmd"
#| include: FALSE
```

Functions like `starts_with()`, `contains()` or `matches()` are
__selection helpers__ that only work in a selection context, e.g.
`dplyr::select()` or the `cols` argument of `tidyr::pivot_longer()`.

Using a selection helper anywhere else results in an error:

```{r, error = TRUE}
starts_with("foo")

mtcars[contains("foo")]

subset(mtcars, select = matches("foo"))
```

If you see this error, you may have used a selection helper in the
wrong place, possibly as the result of a typo (e.g. misplaced comma or
wrong argument name). Alternatively, you may be deliberately trying
to reduce duplication in your code by extracting out a selection into
a variable:

```{r, error = TRUE}
my_vars <- c(name, species, ends_with("color"))
```

To make this work you'll need to do two things:

* Wrap the whole thing in a function
* Use `any_of()` or `all_of()` instead of bare variable names

```{r}
my_vars <- function() {
  c(any_of(c("name", "species")), ends_with("color"))
}
dplyr::select(starwars, my_vars())
```
