---
title: "The tidyselect syntax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{syntax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4
)
options(
  crayon.enabled = FALSE
)
```

This is a technical description of the tidyselect syntax.

```{r setup}
library(tidyselect)
library(magrittr)

# FIXME: `select_pos()` should be exported as `eval_select()`
select_pos <- function(data, ...) {
  tidyselect:::select_pos(data, !!rlang::expr(c(...)))
}

# For better printing
mtcars <- tibble::as_tibble(mtcars)
iris <- tibble::as_tibble(iris)
```


## Sets of variables

The tidyselect syntax is all about __sets__ of variables, internally
represented by integer vectors of __positions__. `c(1L, 2L, 1L)`
represents the set of the first and second variables.

If a vector of positions contains duplicates, they are normally
treated as the same element since they represent sets. An exception to
this occurs with named elements whose names differ. If the names don't
match, they are treated as different elements in order to allow
renaming a variable to multiple names (see section on Renaming
variables).

The syntax of tidyselect is generally designed for set combination.
For instance, `c(foo(), bar())` represents the union of the variables
in `foo()` and those in `bar()`.


### Bare names

Within _data-expressions_ (see Evaluation section), bare names
represent their own positions, i.e. a set of size 1. The following
expressions are equivalent:

```{r}
mtcars %>% select_pos(mpg:hp, -cyl, vs)

mtcars %>% select_pos(1:4, -2, 8)
```


### The `:` operator

`:` can be used to select consecutive variables between two positions.
It returns the corresponding sequence of positions.

```{r}
mtcars %>% select_pos(2:4)
```

Because bare names represent their own positions, it is easy to select
a range of variables:

```{r}
mtcars %>% select_pos(cyl:hp)
```


### Boolean operators

Boolean operators provide a more intuitive approach to set
combination. Though sets are internally represented with vectors of
positions, they could equally be represented with a full logical
vector of inclusion indicators (taking the `which()` of this vector
would then recover the positions). The boolean operators should be
considered in terms of the logical representation of sets.

The `|` operator takes the __union__ of two sets:

```{r}
iris %>% select_pos(starts_with("Sepal") | ends_with("Width"))
```

The `&` operator takes the __intersection__ of two sets:

```{r}
iris %>% select_pos(starts_with("Sepal") & ends_with("Width"))
```

The `!` operator takes the __complement__ of a set:

```{r}
iris %>% select_pos(!ends_with("Width"))
```

Taking the intersection with a complement produces a set
__difference__:

```{r}
iris %>% select_pos(starts_with("Sepal") & !ends_with("Width"))
```


### Dots, `c()`, and unary `-`

tidyselect functions can take dots like `dplyr::select()`, or a named
argument like `tidyr::pivot_longer()`. In the latter case, the dots
syntax is accessible via `c()`. In fact `...` syntax is implemented
through `c(...)` and is thus completely equivalent.

```{r}
mtcars %>% select_pos(mpg, disp:hp)

mtcars %>% select_pos(c(mpg, disp:hp))
```

Dots and `c()` are syntax for:

* Set union or set difference
* Renaming variables

Non-negative inputs are recursively joined with `union()`. The
precedence is left-associative, just like with boolean operators.
These expressions are all syntax for _set union_:

```{r}
iris %>% select_pos(starts_with("Sepal"), ends_with("Width"), Species)

iris %>% select_pos(starts_with("Sepal") | ends_with("Width") | Species)

iris %>% select_pos(union(union(starts_with("Sepal"), ends_with("Width")), 5L))
```

Unary `-` is normally syntax for _set difference_:

```{r}
iris %>% select_pos(starts_with("Sepal"), -ends_with("Width"), -Sepal.Length)

iris %>% select_pos(setdiff(setdiff(starts_with("Sepal"), ends_with("Width")), 1L))
```

If the first `...` or `c()` input is negative, an implicit
`everything()` is appended.

```{r}
iris %>% select_pos(-starts_with("Sepal"))

iris %>% select_pos(everything(), -starts_with("Sepal"))

iris %>% select_pos(setdiff(everything(), starts_with("Sepal")))
```

In this case, unary `-` is syntax for _set complement_. Unary `-` and
`!` are equivalent:

```{r}
iris %>% select_pos(-starts_with("Sepal"))

iris %>% select_pos(!starts_with("Sepal"))
```

Each level of `c()` is independent. In particular, a nested `c()`
starting with `-` always stands for set complement:

```{r}
iris %>% select_pos(c(starts_with("Sepal"), -Sepal.Length))

iris %>% select_pos(c(starts_with("Sepal"), c(-Sepal.Length)))
```

In boolean terms, these expressions are equivalent to:

```{r}
iris %>% select_pos(starts_with("Sepal") & !Sepal.Length)

iris %>% select_pos(starts_with("Sepal") | !Sepal.Length)
```

In general, when unary `-` is used alone outside `...` or `c()`, it
stands for set complement.


### Renaming variables

When named inputs are provided in `...` or `c()`, the selection is
renamed. If the inputs are already named, the outer and inner names
are pasted together with a `...` separator. Otherwise the outer
names is propagated to the selected elements according to the
following rules:

- With data frames, a numeric suffix is appended because columns
  must be uniquely named.

- With normal vectors, the name is simply assigned to all selected
  inputs.

Unnamed elements match any names:

- `a | c(foo = a)` is equivalent to `c(foo = a)`.
- `a & c(foo = a)` is equivalent to `c(foo = a)`.
- `c(foo = a) & c(bar = a)` is equivalent to `c()`.
- `c(foo = a) | c(bar = a)` is equivalent to `c(foo = a, bar = a)`.

Because unnamed elements match any named ones, it is possible to
select multiple elements and rename one of them:

```{r}
iris %>% select_pos(!Species, foo = Sepal.Width)
```


### Predicate functions

Predicate function objects can be supplied as input. They are applied
to all elements of the input, and should return `TRUE` or `FALSE` to
indicate inclusion. Predicates in data-expressions are effectively
expanded to the set of variables that they represent:

```{r}
iris %>% select_pos(is.numeric)

iris %>% select_pos(is.factor)

iris %>% select_pos(is.numeric | is.factor)

iris %>% select_pos(is.numeric & is.factor)
```


## Selection helpers

We call _selection helpers_ any function that inspects the currently
active variables with `peek_vars()` and returns a selection.

- `peek_vars()` returns a character vector of names.
- The returned selection can be any output conforming to the types
  described in the Data types section.

Examples of selection helpers are `all_of()`, `contains()`, or
`last_col()`. These selection helpers are evaluated as env-expressions
(see Evaluation section).


## Supported data types

The following data types can be returned from selection helpers or
forced via `!!` or `force()` (the latter works in tidyselect because
it is treated as an env-expression, see Evaluation section):

- Vectors of positions:

    ```{r}
    iris %>% select_pos(force(c(1, 3)))
    ```

- Vectors of names. These are matched and transformed to positions.

    ```{r}
    iris %>% select_pos(force(c("Sepal.Length", "Petal.Length")))
    ```

- Predicate functions. These are applied to all elements to determine
  inclusion.

    ```{r}
    iris %>% select_pos(force(is.numeric))
    ```


## Evaluation

### Data-expressions and env-expressions

tidyselect is not a typical tidy evaluation UI. The main difference is
that there is no data masking. In a typical tidy eval function,
expressions are evaluated with data-vars first in scope, followed by
env-vars:

```{r}
mask <- function(data, expr) {
  rlang::eval_tidy(rlang::enquo(expr), data)
}

foo <- 10
cyl <- 200

# `cyl` represents the data frame column here:
mtcars %>% mask(cyl * foo)
```

It is possible to bypass the data frame variables by forcing symbols
to be looked up in the environment with `!!` or `.env`:

```{r}
mtcars %>% mask(!!cyl * foo)
mtcars %>% mask(.env$cyl * foo)
```

With tidyselect, there is no such hierarchical data masking. Instead,
expressions are evaluated either in the context of the data frame or
in the user environment, without overlap. The scope of lookup depends
on the kind of expression:

1.  __data-expressions__ are evaluated in the data frame only. This
    includes bare symbols, the boolean operators, `-`, `:`, and `c()`.
    You can't refer to environment-variables in a data-expression:

    ```{r, error = TRUE}
    cyl_pos <- 2
    mtcars %>% select_pos(mpg | cyl_pos)
    ```

2.  __env-expressions__ are evaluated in the environment. This
    includes all calls other than those mentioned above, as well as
    symbols that are part of those calls. You can't refer to
    data-variables in a data-expression:

    ```{r, error = TRUE}
    mtcars %>% select_pos(all_of(mpg))
    ```

Because the scoping is unambiguous, you can safely refer to env-vars
in an env-expression, without having to worry about potential naming
clashes with data-vars:

```{r}
x <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# `ncol(x)` is an env-expression, so `x` represents the data frame in
# the environment rather than the column in the data frame
x %>% select_pos(2:ncol(x))
```

If you have variable names in a character vector, it is safe to refer
to the env-var containing the names with `all_of()` because it is an
env-expression:

```{r}
y <- c("y", "z")
x %>% select_pos(all_of(y))
```

Note that currently, env-vars are still allowed in some
data-expressions, for compatibility. However this is in the process of
being deprecated and you should see a note recommending to use
`all_of()` instead. This note will become a deprecation warning in the
future, and then an error.

```{r}
mtcars %>% select_pos(cyl_pos)
```


### Predicate functions within data-expressions

There is one exception to the unambiguous lookup rules in data
expressions. If a data-var is not found in the data frame, tidyselect
tries to find a predicate function in the environment.

This introduces an ambiguity when a predicate function has the same
name as a column. This is a conscious design decision:

- It makes it easier to use predicate functions interactively.
- Predicate functions and data frame columns should normally have
  different names.

If you need 100% unambiguity, use `all_of()` to refer to data frame
columns and force predicate functions with `!!` or
`force()`. `all_of()` will fail if the data frame doesn't contain the
columns, even if a function of the same name is found in the
environment:

```{r, error = TRUE}
Species <- function(x) is.numeric(x)

# Since `Species` is defined as a function, you won't get an error
# even though the data-var doesn't exist:
mtcars %>% select_pos(Species)

# Use `all_of()` to be explicit that you want a data-var, not an env-var:
mtcars %>% select_pos(all_of("Species"))
```

Similarly, you can be explicit about finding predicate functions in
the environment by forcing the symbols:

```{r}
# By default the data-var has precedence over the env-var:
iris %>% select_pos(Species)

# Use the force operator `!!` to force the env-var `Species`:
iris %>% select_pos(!!Species)

# Use the env-expression `force()` to force the env-var `Species`:
iris %>% select_pos(force(Species))
```


### Arithmetic operators

Within data-expressions (see Evaluation section), `+`, `*` and `/` are
overridden to cause an error. This is to prevent confusion stemming
from normal data masking usage where variables can be transformed on
the fly:

```{r, error = TRUE}
mtcars %>% select_pos(cyl^2)

mtcars %>% select_pos(mpg * wt)
```
